Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-04-23 Paper/manuscript_v2/manuscript_07_01.tex
layout the paper

\begin{figure}
\centering
%\subfigure{\includegraphics[scale=0.38, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig1_legend_01.pdf}}\\
%\vspace{-9mm}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 10mm 0mm, clip]{ch5_fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig1_Sigma_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 3mm 0mm, clip]{ch5_fig1_Cov_J.pdf}}
%\subfigure{\includegraphics[scale=0.38, trim = 29mm 48mm 124mm 10mm, clip]{ch5_fig1_legend.pdf}}
\caption{Mean posterior predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the exact GP model (dashed red line) and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 10mm 0mm, clip]{ch5_fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig2_Sigma_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 3mm 0mm, clip]{ch5_fig2_Cov_L.pdf}}
%\subfigure{\includegraphics[scale=0.38, trim = 29mm 48mm 124mm 10mm, clip]{ch5_fig2_legend.pdf}}
\caption{Mean posterior predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the exact GP model (dashed red line) and the HSGP model for different values of the boundary factor $c$, with a large enough fixed number of basis functions.}
  \label{fig2_Post_L}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig4_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig4_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the exact GP model. RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$ (left). RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$ (right). }
  \label{fig4_MSE_vs_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig5_magnitud_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 28mm 119mm 0mm, clip]{ch5_fig5_legend.pdf}}
\caption{Estimated lengthscale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.42, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Mean posterior predictions of both exact GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic lengthscales ($\ell$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{true function}). %The dashed red line of the exact GP can hardly be seen in the plots because it is almost perfectly overlayed by the blue lines.
}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

\begin{figure}
\begin{flushleft}
\begin{tabular}{ccc}
\multicolumn{3}{c}{ \includegraphics[scale=0.42, trim = 25mm 117mm 90mm 45mm, clip]{ch5_fig8_legend.pdf}}\\
\includegraphics[scale=0.31, trim = 0mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part1.pdf} & \hspace{-4mm}
\includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part2.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part3.pdf} \\ 
\includegraphics[scale=0.31, trim = 0mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part4.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part5.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part6.pdf}
\end{tabular}
\end{flushleft}
\caption{Data-generating functional lengthscales ($\ell$), of the various datasets illustrated in Figure \ref{fig7_posterior_varing_lscale_part1}, versus the corresponding lengthscale estimates $(\hat{\ell})$ from the exact GP and HSGP models. 95\% confident intervals of the lengthscale estimates are plotted as dot lines. The different plots represent the use of different number of basis functions $m$ in the HSGP model. The dashed black line represents the recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c} that can be closely approximated by the HSGP model in every case.}
  \label{fig8_Tlscale_vs_Elscale}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.38]{ch5_fig9_MSE_varing_lscale.pdf}
\caption{RMSE of the HSGP models with different number of basis functions $m$, for the various datasets with different wiggly effects ($\ell$).}
  \label{fig9_MSE_varing_lscale}
\end{figure}

@@

\section{Case studies}\label{ch5_sec_cases}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig11_MSE_exI_inter.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig11_MSE_exI_extra.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 29mm 55mm 110mm 0mm, clip]{ch5_fig11_MSE_exI_legend.pdf}}
\caption{Standardized root mean square error (SRMSE) of the different methods against the data-generating function. SRMSE for interpolation (left) and SRMSE for extrapolation (right). The standard deviation of the mean of the SRMSE is plotted as dashed lines.}
  \label{ch5_fig11_MSE_exI_inter}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig11_time_exI.pdf}}
\caption{Computational time (y-axis), in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The y-axis is on a logarithmic scale. The standard deviation of the computational time is plotted as dashed lines.}
  \label{ch5_fig11_time_exI}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig27_posteriors_oneyear_birthday.pdf}}
\caption{Posterior means of the function $\mu(\cdot)$ for the year 1972 of the series. The special days effects pattern ($f_4(\cdot)$) in the year is also represented, as well as the long-term trend ($f_1(\cdot)$) and year effects pattern ($f_2(\cdot)$). }
  \label{ch5_fig27_posteriors_oneyear_birthday}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig27_posteriors_onemonth_birthday.pdf}}
\caption{Posterior means of the function $\mu(\cdot)$ for the month of January of 1972. The week effects pattern ($f_3(\cdot)$) in the month is also represented, as well as the long-term trend ($f_1(\cdot)$), year effects pattern ($f_2(\cdot)$) and special days effects pattern ($f_4(\cdot)$). }
  \label{ch5_fig27_posteriors_onemonth_birthday}
\end{figure}

\subsection{Leukemia data}\label{ch5_sec_bf_caseVI}

@@

\begin{figure}
\centering
\includegraphics[scale=0.70, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig21_posteriors_leukemia_02.pdf}
\caption{Expected lifetime conditional comparison for each predictor with other predictors fixed to their mean values. The thick line in each graph is the posterior mean estimated using a HSGP model, and the thin lines represent pointwise 95\% credible intervals.}
  \label{ch5_fig21_posteriors_leukemia}
\end{figure}

@@

Figure \ref{ch5_fig21_posteriors_leukemia} shows estimated conditional functions of each predictor with all others fixed to their mean values. These posterior estimates correspond to the HSGP model with $m=10$ basis functions and $c=3$ boundary factor. There are clear  non-linear patterns and the right bottom subplot also shows that the conditional function associated with WBC has an interaction with TDI.
Figure~\ref{ch5_fig22_elpd_leukemia} shows the expected log predictive density \citep[ELPD; ][]{vehtari_2012,Vehtari+Gelman+Gabry:2017_practical} and time of computation as function of the number of univariate basis functions $m$ ($m^{\ast}=m^D$ in equation (\ref{eq_approxf_multi})) and boundary factor $c$. As the functions are smooth, a few number of basis functions and a large boundary factor are required to obtain a good approximation (Figure \ref{ch5_fig22_elpd_leukemia}-left); Small boundary factors are not appropriate for models for large lengthscales, as can be seen in Figure \ref{fig6_lscale_vs_J_vs_c}. Increasing the boundary factor also significantly increases the time of computation (Figure \ref{ch5_fig22_elpd_leukemia}-right). With a moderate number of univariate basis functions ($m=15$), the HSGP model becomes slower than the exact GP model, in this specific application with $3$ input variables, as the total number of multivariate basis functions becomes $15^3 = 3375$ and is therefore quite high. 
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig22_elpd_leukemia.pdf}} 
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig23_time_leukemia.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 45mm 120mm 0mm, clip]{ch5_fig24_legend_leukemia.pdf}}
\caption{Expected log predictive density (ELPD; left) and time of computation in seconds per iteration (iteration of the HMC sampling method; right) as a function of the number of basis functions $m$ and boundary factor $c$.}
  \label{ch5_fig22_elpd_leukemia}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig28_m_lscale_periodic_appendix.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig28_m_lscale_periodic_appendix_zoom.pdf}}
\caption{Relation among the minimum number of terms $J$ in the approximation and the lengthscale ($\ell$) of the periodic squared exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{ch5_fig28_m_lscale_periodic}
\end{figure}

The function values of a GP model with this low-rank representation of the periodic exponential covariance function can be easily derived. Considering the identity
%
\begin{equation*}
\cos(j\omega_0 (x-x'))=\cos(j\omega_0 x) \cos(j\omega_0 x') + \sin(j\omega_0 x) \sin(j\omega_0 x'),
\end{equation*}
the covariance $k(\tau)$ in equation (\ref{ch5_eq_cov_periodic_taylor_approx}) can be written as
%
\begin{equation} \label{ch5_eq_cov_periodic_taylor_approx_2}
k(x,x')= \alpha \left( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \right).
\end{equation}
With this approximation for the periodic squared exponential covariance function $k(x,x')$, the approximate GP model $f(x) \sim \GP\left(0,k(x,x')\right)$ equivalently leads to a linear representation of $f(\cdot)$ via
%
\begin{equation} \label{ch5_eq_f_period}
f(x) \approx \alpha^{1/2} \left( \sum_{j=0}^J  \tilde{q}_j \cos(j\omega_0 x) \beta_j +  \sum_{j=1}^J \tilde{q}_j \sin(j\omega_0 x) \beta_{J+1+j} \right),
\end{equation}
where $\beta_j \sim \Normal(0,1)$, with $j=1,\dots,2J+1$. The cosine $\cos(j\omega_0 x)$ and sinus $\sin(j\omega_0 x)$ terms do not depend on the covariance hyperparameters $\ell$. The only dependence on the hyperparameter $\ell$ is through the coefficients $\tilde{q}_j$, which are $J$-dimensional. The computational cost of this approximation scales as $O\big(n(2J+1) + (2J+1)\big)$, where $n$ is the number of observations and $J$ the number of cosine terms.
The parameterization in equation (\ref{ch5_eq_f_period}) is naturally in the non-centered form with independent prior distributions on
$\beta_j$, which makes posterior inference easier.

>>>>

\begin{figure}[!t]
\centering
%\subfigure{\includegraphics[scale=0.38, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig1_legend_01.pdf}}\\
%\vspace{-9mm}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 10mm 0mm, clip]{ch5_fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 5mm 10mm 0mm, clip]{ch5_fig1_Sigma_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{ch5_fig1_Cov_J.pdf}}
%\subfigure{\includegraphics[scale=0.38, trim = 29mm 48mm 124mm 10mm, clip]{ch5_fig1_legend.pdf}}
\vspace{-1mm}
\caption{Mean posterior predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the exact GP model (dashed red line) and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 10mm 0mm, clip]{ch5_fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 5mm 10mm 0mm, clip]{ch5_fig2_Sigma_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{ch5_fig2_Cov_L.pdf}}
%\subfigure{\includegraphics[scale=0.38, trim = 29mm 48mm 124mm 10mm, clip]{ch5_fig2_legend.pdf}}
\vspace{-1mm}
\caption{Mean posterior predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the exact GP model (dashed red line) and the HSGP model for different values of the boundary factor $c$, with a large enough fixed number of basis functions.}
  \label{fig2_Post_L}
\end{figure}

@@

\begin{figure}[!t]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{ch5_fig4_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 10mm 0mm, clip]{ch5_fig4_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the exact GP model. RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$ (left). RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$ (right). }
  \label{fig4_MSE_vs_J}
\end{figure}

\begin{figure}[!h]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 5mm 0mm, clip]{ch5_fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 10mm 0mm, clip]{ch5_fig5_magnitud_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 35mm 119mm 0mm, clip]{ch5_fig5_legend.pdf}}
\caption{Estimated lengthscale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure}

@@

@@

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.42, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part9.pdf}}\\
\caption{Mean posterior predictions of both exact GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic lengthscales ($\ell$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{true function}). %The dashed red line of the exact GP can hardly be seen in the plots because it is almost perfectly overlayed by the blue lines.
}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

\begin{figure}
\begin{flushleft}
\begin{tabular}{ccc}
\multicolumn{3}{c}{ \includegraphics[scale=0.42, trim = 25mm 117mm 90mm 45mm, clip]{ch5_fig8_legend.pdf}}\\
\includegraphics[scale=0.31, trim = 0mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part1.pdf} & \hspace{-4mm}
\includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part2.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part3.pdf} \\ 
\includegraphics[scale=0.31, trim = 0mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part4.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part5.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part6.pdf}
\end{tabular}
\end{flushleft}
\caption{Data-generating functional lengthscales ($\ell$), of the various datasets illustrated in Figure \ref{fig7_posterior_varing_lscale_part1}, versus the corresponding lengthscale estimates $(\hat{\ell})$ from the exact GP and HSGP models. 95\% confident intervals of the lengthscale estimates are plotted as dot lines. The different plots represent the use of different number of basis functions $m$ in the HSGP model. The dashed black line represents the recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c} that can be closely approximated by the HSGP model in every case.}
  \label{fig8_Tlscale_vs_Elscale}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.38, trim = 0mm 9mm 0mm 0mm, clip]{ch5_fig9_MSE_varing_lscale.pdf}
\caption{RMSE of the HSGP models with different number of basis functions $m$, for the various datasets with different wiggly effects ($\ell$).}
  \label{fig9_MSE_varing_lscale}
\end{figure}

\newpage
\section{Case studies}\label{ch5_sec_cases}

@@

\begin{figure}[!t]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 5mm 5mm 0mm, clip]{ch5_fig11_MSE_exI_inter.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 5mm 10mm 0mm, clip]{ch5_fig11_MSE_exI_extra.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 29mm 55mm 110mm 0mm, clip]{ch5_fig11_MSE_exI_legend.pdf}}
\caption{Standardized root mean square error (SRMSE) of the different methods against the data-generating function. SRMSE for interpolation (left) and SRMSE for extrapolation (right). The standard deviation of the mean of the SRMSE is plotted as dashed lines.}
  \label{ch5_fig11_MSE_exI_inter}
\end{figure}

\begin{figure}[!h]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 5mm 10mm 0mm, clip]{ch5_fig11_time_exI.pdf}}
\caption{Computational time (y-axis), in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The y-axis is on a logarithmic scale. The standard deviation of the computational time is plotted as dashed lines.}
  \label{ch5_fig11_time_exI}
\end{figure}

@@

\begin{figure}[!t]
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig27_posteriors_oneyear_birthday.pdf}}
\caption{Posterior means of the function $\mu(\cdot)$ for the year 1972 of the series. The special days effects pattern ($f_4(\cdot)$) in the year is also represented, as well as the long-term trend ($f_1(\cdot)$) and year effects pattern ($f_2(\cdot)$). }
  \label{ch5_fig27_posteriors_oneyear_birthday}
\end{figure}

\begin{figure}[!h]
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig27_posteriors_onemonth_birthday.pdf}}
\caption{Posterior means of the function $\mu(\cdot)$ for the month of January of 1972. The week effects pattern ($f_3(\cdot)$) in the month is also represented, as well as the long-term trend ($f_1(\cdot)$), year effects pattern ($f_2(\cdot)$) and special days effects pattern ($f_4(\cdot)$). }
  \label{ch5_fig27_posteriors_onemonth_birthday}
\end{figure}

\newpage 
\subsection{Leukemia data}\label{ch5_sec_bf_caseVI}

@@

\begin{figure}
\centering
\includegraphics[scale=0.70, trim = 0mm 5mm 0mm 0mm, clip]{ch5_fig21_posteriors_leukemia_02.pdf}
\caption{Expected lifetime conditional comparison for each predictor with other predictors fixed to their mean values. The thick line in each graph is the posterior mean estimated using a HSGP model, and the thin lines represent pointwise 95\% credible intervals.}
  \label{ch5_fig21_posteriors_leukemia}
\end{figure}

@@

Figure \ref{ch5_fig21_posteriors_leukemia} shows estimated conditional functions of each predictor with all others fixed to their mean values. These posterior estimates correspond to the HSGP model with $m=10$ basis functions and $c=3$ boundary factor. There are clear  non-linear patterns and the right bottom subplot also shows that the conditional function associated with WBC has an interaction with TDI.
Figure~\ref{ch5_fig22_elpd_leukemia} shows the expected log predictive density \citep[ELPD; ][]{vehtari_2012,Vehtari+Gelman+Gabry:2017_practical} and time of computation as function of the number of univariate basis functions $m$ ($m^{\ast}=m^D$ in equation (\ref{eq_approxf_multi})) and boundary factor $c$. As the functions are smooth, a few number of basis functions and a large boundary factor are required to obtain a good approximation (Figure \ref{ch5_fig22_elpd_leukemia}-left); Small boundary factors are not appropriate for models for large lengthscales, as can be seen in Figure \ref{fig6_lscale_vs_J_vs_c}. Increasing the boundary factor also significantly increases the time of computation (Figure \ref{ch5_fig22_elpd_leukemia}-right). With a moderate number of univariate basis functions ($m=15$), the HSGP model becomes slower than the exact GP model, in this specific application with $3$ input variables, as the total number of multivariate basis functions becomes $15^3 = 3375$ and is therefore quite high. 
%
\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{ch5_fig22_elpd_leukemia.pdf}} 
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 10mm 0mm, clip]{ch5_fig23_time_leukemia.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 50mm 120mm 0mm, clip]{ch5_fig24_legend_leukemia.pdf}}
\caption{Expected log predictive density (ELPD; left) and time of computation in seconds per iteration (iteration of the HMC sampling method; right) as a function of the number of basis functions $m$ and boundary factor $c$.}
  \label{ch5_fig22_elpd_leukemia}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 8mm 5mm 0mm, clip]{ch5_fig28_m_lscale_periodic_appendix.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 8mm 5mm 0mm, clip]{ch5_fig28_m_lscale_periodic_appendix_zoom.pdf}}
\caption{Relation among the minimum number of terms $J$ in the approximation and the lengthscale ($\ell$) of the periodic squared exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{ch5_fig28_m_lscale_periodic}
\end{figure}

The function values of a GP model with this low-rank representation of the periodic exponential covariance function can be easily derived. Considering the identity
%
\begin{equation*}
\cos(j\omega_0 (x-x'))=\cos(j\omega_0 x) \cos(j\omega_0 x') + \sin(j\omega_0 x) \sin(j\omega_0 x'),
\end{equation*}
the covariance $k(\tau)$ in equation (\ref{ch5_eq_cov_periodic_taylor_approx}) can be written as
%
\begin{equation} \label{ch5_eq_cov_periodic_taylor_approx_2}
k(x,x')\approx \alpha \left( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \right).
\end{equation}
With this approximation for the periodic squared exponential covariance function $k(x,x')$, the approximate GP model $f(x) \sim \GP\left(0,k(x,x')\right)$ equivalently leads to a linear representation of $f(\cdot)$ via
%
\begin{equation} \label{ch5_eq_f_period}
f(x) \approx \alpha^{1/2} \left( \sum_{j=0}^J  \tilde{q}_j \cos(j\omega_0 x) \beta_j +  \sum_{j=1}^J \tilde{q}_j \sin(j\omega_0 x) \beta_{J+1+j} \right),
\end{equation}
where $\beta_j \sim \Normal(0,1)$, with $j=1,\dots,2J+1$. The cosine $\cos(j\omega_0 x)$ and sinus $\sin(j\omega_0 x)$ terms do not depend on the covariance hyperparameters $\ell$. The only dependence on the hyperparameter $\ell$ is through the coefficients $\tilde{q}_j$, which are $J$-dimensional. The computational cost of this approximation scales as $O\big(n(2J+1) + (2J+1)\big)$, where $n$ is the number of observations and $J$ the number of cosine terms.
The parameterization in equation (\ref{ch5_eq_f_period}) is naturally in the non-centered form with independent prior distributions on
$\beta_j$, which makes posterior inference easier.
